Digital video platforms and phylogenetic modelling reveal parrot tool use is not rare

### 2024-05-20 manuscript.Rmd
update manuscript and prepare for submission

---
title             : "Crowdsourcing and phylogenetic modelling reveal parrot tool use is not rare"
shorttitle        : "Parrot tool use"

@@

abstract: |
 Putatively rare behaviours, such as tool use, are challenging to study because absence of evidence can arise either from a species’ inability to produce the behaviour or from insufficient research effort. Here, we tackle this challenge by combining crowdsourcing and phylogenetic modelling to estimate actual rates of tool use in parrots. Crowdsourcing on a social media platform revealed novel instances of tool use in 17 parrot species, more than doubling the confirmed number of tool-using parrot species from 11 (3%) to 28 (7%). Phylogenetic modelling ranked additional species that are most likely to be unobserved tool users, suggesting that between 11% and 17% of extant parrot species may be tool users. These discoveries have implications for inferences about the evolutionary drivers and origins of tool use in parrots, revealing associations with relative brain size and feeding generalism and indicating several genera where tool use was likely an ancestral trait. Overall, our findings challenge the assumption that current sampling effort captures the full distribution of putatively rare animal behaviours. Combining our sampling and analysis methods offers a fruitful approach for investigating the distribution, drivers, and origins of other rare behaviours.

keywords          : "tool use; technical intelligence; parrots; crowdsourcing; phylogenetic modelling"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright † These authors contributed equally to this work \par ‡ These authors contributed equally to this work \par * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

\linenumbers

@@

One potentially powerful method for quantifying actual rates of rare animal 
behaviours is crowdsourcing [@Nelson2013]. In a crowdsourcing study, researchers 
collect reports from the general public and/or collate and analyse videos posted 
on social media platforms. This citizen science approach has been widely used in 
ecology to monitor the distributional patterns of species [@Dickinson2010], but 
has also recently been used to uncover a variety of rare animal behaviours, 
including interspecies play in dogs [@Nelson2013], novel problem-solving 
behaviours in horses [@Krueger2019], and socially-learned foraging innovations 
in cockatoos [@Klump2021;@Klump2022]. By casting the net wider than the 
scientific literature, the crowdsourcing method can provide an indication of the 
tool-using species that the literature might be missing.

Even after using this crowdsourcing approach, some tool-users could _still_ 
remain unobserved. One principled framework for identifying these unobserved species 
is to specify a causal model of the process that generates the observed data. We 
propose one such causal model in Figure \@ref(fig:plotDAG). In this model, we 
assume that the presence of tool use in the scientific literature (or in 
crowdsourced reports) is caused by both unobserved tool use capabilities and the 
number of published studies (or the number of crowdsourced reports) for any 
given species. Tool users are more likely to be observed if they are well 
studied, but understudied tool users may go undetected. Furthermore, based on 
existing theories of the evolution of tool use [@Ducatez2015;
@HenkeVonDerMalsburg2020;@Lefebvre1997;@Lefebvre2002;@Lefebvre2004;
@MettkeHofmann2014;@Overington2011;@Sol2005;@Sol2009], we propose that the 
unobserved tool use capabilities are additionally caused by relative brain size, 
feeding strategy, and shared phylogenetic ancestry. Expressing this causal model 
as a statistical model can suggest further species which are likely to be 
unobserved tool-users and, simultaneously, test existing theories of the 
evolutionary drivers of tool use without incorrectly assuming that absence of 
evidence is evidence of absence.

(ref:plotDAGCaption) _Causal model of observed tool use._ Directed acyclic graph 
of the causal relationships between observed tool use and other variables. 
Available scientific data on tool use is caused both by unobserved tool use 
presence and scientific research effort (i.e., number of publications). 
Available crowdsourced data on tool use is caused both by unobserved tool use 
presence and crowdsourcing effort (i.e., number of crowdsourced reports). 
According to theory, unobserved tool use presence should be caused by relative 
brain size (encephalisation quotient) and feeding strategy (generalist vs. 
specialist). These variables all share unobserved confounds generated by shared 
phylogenetic history. Grey circles indicate unobserved variables.

@@

Here, we apply these crowdsourcing and phylogenetic modelling approaches to tool 
use in the parrot order. We focus on tool use in parrots for a number of 
reasons. First, the scientific literature suggests that only a small proportion 
of extant parrot species (11 out of 398; 3%) use tools 
[@Auersperg2009;@Auersperg2010;@Auersperg2011a;@Auersperg2011b;@Auersperg2012;
@Bastos2021;@BentleyCondit2010;@Borsari2005;@Goodman2018;@Heinsohn2017;
@Janzen1976;@Lambert2015;@Lefebvre2002;@OHara2021]. Parrot tool use 
thus provides an ideal test case for examining how robust sampling is in the 
scientific literature. Second, parrots are highly popular as pets. Over 70% of 
all extant parrot species are bred in the aviculture industry and kept as pets 
worldwide [@Anderson2003;@Carrete2008;@Drews2001;@Kelly2014;@Li2014;@Su2015;
@Young2012], enabling us to leverage the power of crowdsourcing on a social 
media platform to search for evidence of tool use [@Nelson2013]. Third, detailed 
data on relative brain sizes [@Flammer2001;@Iwaniuk2005;@Mazengenya2018;
@Olkowicz2016;@Schuck2008;@Silva2017], feeding strategies [@Wilman2016], and 
shared ancestry [@Jetz2012] exist for parrots, allowing us to fit the 
statistical model implied by Figure \@ref(fig:plotDAG) to the entire parrot 
order.

We first present the results from our crowdsourcing survey, in which we collated 
videos of tool use in parrots from an online video platform. This survey reveals 
a number of previously unidentified tool-using parrot species, which we map onto 
the phylogeny of the parrot order. We then describe our statistical model in 
more detail, and use it to (_i_) rank further parrot species that are likely 
unobserved tool users and (_ii_) re-examine key hypotheses regarding the 
evolutionary drivers and origins of tool use in parrots.

@@

## Crowdsourcing reveals tool use in additional parrot species

We surveyed the social media platform YouTube for video evidence of tool use in 
parrots (see Methods for detailed search criteria). In our search, we used the 
standard criteria for identifying tool use in the literature, defining "true" 
tool use behaviour as the manipulation of an unattached object as an extension 
of the animal's body to achieve a goal [@Shumaker2011], while "borderline" tool 
use involved the use of an object that was still attached to a substrate
[@Seed2010].

@@

(ref:plotPhylo1Caption) _Results of crowdsourcing video survey and phylogenetic 
survival cure modelling mapped onto a maximum clade credibility phylogeny of the 
parrot order._ Orange points in the inner ring indicate species observed in the 
video survey, with point size scaled by the number of videos for each species 
(note that three species observed only in the video survey are not present in 
the phylogeny due to a lack of genomic data: _Psittacara erythrogenys_, 
_Psittacus timneh_, and _Aratinga nenday_). Blue points in the outer ring 
indicate species observed in the scientific literature, with point size scaled 
by the number of papers published on each species. Yellow species names indicate 
the top ten most likely tool-using species from our phylogenetic survival cure 
model which were not observed in the scientific literature or the video survey. 
Total _n_ = `r nrow(tar_read(d))` species.

@@

In addition to the 28 tool-using species identified in the literature and our 
video survey, we fitted a Bayesian phylogenetic survival cure model to rank 
further species that are likely to be undetected tool-users (i.e., tool-using 
species with no tool use reported in the literature or in crowdsourced videos).

@@

Our tool use problem has the same features. We are modelling a time-to-event; 
specifically, the amount of "time" (i.e., observation opportunities measured as 
the number of published papers or crowdsourced videos) until tool use is 
identified. This is right-censored data because many species will not have had 
tool use identified when we measure them. Moreover, we can assume that a certain 
proportion of the population will never experience the event -- that is, they 
are a non-tool-user and so we will never identify tool use no matter how long we 
measure them for.

@@

Since the earliest anecdotes of parrots using tools by Wallace in the 1880s
[@Wallace1869] and more systematic anecdotal reports in the 1970s [@Goodall1971], 
only 11 parrot species (3% of all extant parrots) have been documented as tool 
users in the scientific literature. Our study used crowdsourcing to identify 17 
additional tool-using parrot species that are new to science, more than doubling 
the overall count to 28 species (7%). These species consisted of cockatoos 
(_Cacatuidae_), Old World parrots (_Psittacinae_), and neotropical parrots 
(_Arinae_). Beyond the crowdsourcing survey, the strong phylogenetic signal in 
our dataset allowed us to use phylogenetic information, along with other 
variables, to infer the unobserved probabilities of tool use across the parrot 
order. Our phylogenetic survival cure model incorporated information on 
phylogenetic history, research effort, relative brain size, and feeding 
specialisation to rank parrot species that were most likely to be undetected 
tool users. The sum of probabilities from this model implied that between 15 
and 41 of the species without observed evidence of tool use are likely to be 
undetected tool users, suggesting that the true proportion of tool users may be
as high as 17%.

@@

All instances of tool use in our crowdsourcing survey met the established
criteria for tool use in the literature: the manipulation of an unattached 
object as an extension of the animal's body to achieve a goal [@Shumaker2011].
However, a striking feature of the dataset is that the 17 additional species
identified in the video survey were exclusively observed using tools for
self-scratching. While parrots in the wild use tools to achieve a variety of 
goals, including extractive foraging and preening [@Goodman2018;@OHara2021],
parrots in captivity only require tools to achieve the latter goal. Despite this
difference in the type of tool use observed, self-scratching is nonetheless a 
major category of animal tool use. Self-scratching also meets the definition for
a more complex type of embodied tool use known as 'tooling', the deliberate 
generation of a mechanical interface by using an object to manipulate another 
target or surface [@Fragaszy2018]. This makes it difficult to dismiss
self-scratching as unimportant: to do so would require a redefinition of
tool-use in non-human animals. Moreover, the fact that we found strong 
phylogenetic clustering of self-scratching tool use from the video survey with 
other examples of tool use from the literature supports a common underlying 
cognitive mechanism. In line with this, some of the species of parrots that use
tools for self-scratching in captivity also use tools for other purposes in the 
wild [@Auersperg2012;@Bastos2021;@Goodman2018;@OHara2021].

There was little evidence to suggest that the observations of self-scratching
tool use in the video survey were merely unintentional accidents or explicitly 
trained behaviours. Individual parrots often used tools slowly and repetitively 
over long periods of time, even across multiple different videos, suggesting 
that their behaviour was not random or accidental [@Sandor2020]. Parrots 
preferentially employed self-scratching tools on areas of their body that were 
otherwise inaccessible, with 96% of all instances involving scratching of the 
head or neck, suggesting intentional tool use. For 60% of the species in the 
video survey, we found two or more videos of repetitive and sustained scratching 
by different individuals of the same species in separate households, suggesting 
that the manipulations were intentional and recurring events that did not 
represent unusual stereotypies of any single individual. Regarding the 
possibility of training or cueing from owners, over half of the videos contained 
no evidence of human interaction aside from filming the behaviour. Humans only 
handed parrots their tools in two of the videos, and none of the videos featured 
owners directly rewarding tool use behaviours with food. Finally, the high 
levels of phylogenetic signal in our data provide strong evidence that the 
observations from our video survey reflect biologically-endowed capacities for 
tool use rather than accidental or trained behaviours, which would likely appear 
uniformly across the phylogeny.

While missing data imputation is becoming more common in phylogenetic 
analyses [@Debastiani2021], the important distinction between absence of 
evidence and evidence of absence has not been given as much attention. Our 
phylogenetic analysis provides one approach to this problem by distinguishing 
between true absences of tool use and absences of tool use due to a lack of 
research effort in the scientific literature or in crowdsourced videos. To 
achieve this, we explicitly modelled the measurement of the outcome variable 
along a research effort time series, such that species with lower research 
effort in the literature or in videos were likely to be censored. In line with 
our causal model, we also included relative brain size, feeding strategy, and 
phylogenetic history as predictors of unobserved tool use. We encourage 
researchers to test this model by directing future study efforts towards the 
parrot species with the highest probabilities of being undetected tool-users. 
Future work should also refine the causal model in Figure \@ref(fig:plotDAG) to 
provide more certain estimates of tool use probabilities, either by including 
additional predictor variables or modelling further causes of measurement error 
in the taxonomic record, such as species abundance and geographic 
accessibility [@Ducatez2014].

In conclusion, we have shown that the scientific literature has insufficiently 
captured the full distribution of tool use in the parrot order. Our 
crowdsourcing survey more than doubled the number of known tool using parrot 
species from 11 to 28, and our phylogenetic model suggested that the true 
proportion of parrot tool users could be as high as 17% of all species in this 
order. These discoveries have implications for theories of the evolutionary 
drivers and origins of tool use in parrots. Beyond parrot tool use, the 
crowdsourcing and phylogenetic methods used in this study have the potential to 
be applied to other rarely observed behaviours, including tool use in other 
taxa [@Hunt2013], rhythmic entrainment in birds [@Benichov2016;@Patel2009;
@tenCate2016;@Wilson2016], teasing behaviours in primates [@Eckert2020], and 
tactical deception across all animals [@BroJorgensen2010;@Byrne1988;@Byrne1991]. 
We hope that these methods will continue to uncover a diverse array of ephemeral 
behaviours that have as yet gone undetected in the scientific literature.

@@

Our video search was conducted on YouTube in July 2020. Search terms included 
"parrot using tool" and variants (e.g., "macaw using tool", "lorikeet using 
tool", "parakeet using tool"), "tool use in parrot", "parrot tool use", "parrot 
scratching itself" (included after we found several videos demonstrating 
self-care tool use in previous searches) and equivalent terms (e.g., "parrot 
preening itself", "parrot grooming itself", "parrot scratching"). For all 
species that did not display results including object manipulation or scratching 
behaviours, we also searched the species’ common name(s) + "tool use", as well 
as the species’ common name(s) + "scratching". We also searched for translations 
of the terms "parrot tool use" and "parrot scratching" in languages for all 
countries where bird ownership was reported as >5% [@GlobalGfkSurvey], namely, 
Turkish, Czech, Polish, French, Italian, Dutch, German, Russian, Spanish, 
Portuguese, and Mandarin.

@@

@@

For modelling purposes, we coded research effort in both the scientific 
literature and the crowdsourced videos. For the scientific literature, we 
operationalised research effort as the number of papers published for each 
species' Latin name up to and including the first paper containing tool use for 
that species. If no tool use had been identified in the scientific literature 
for a species, then we coded the total number of papers published to date. We 
used the scientific database Scopus for coding the number of published papers. 
For the crowdsourced videos, we coded research effort as the number of search 
hits for each species on YouTube. If tool use had been identified on YouTube, we 
estimated the number of search hits when the first video of tool use was 
published on YouTube, assuming linear growth of search hits since the inception 
of YouTube. If tool use had not been identified, we used the current number of 
search hits.

@@

To infer unobserved probabilities of tool use across parrots, we proposed a 
causal model of observed tool use (Figure \@ref(fig:plotDAG)). We assumed that 
observed tool use in the scientific literature and in the crowdsourced videos is 
caused by both the unobserved presence or absence of tool use and research 
effort, proxied by the number of papers published on a species and the number of 
videos published on a species. Tool users are more likely to be observed if they 
are well studied, but understudied tool users may go undetected. In addition, 
based on theory, we also assumed that unobserved tool use is caused by feeding 
strategy and relative brain size [@Lefebvre2004;@Lefebvre1997;@Lefebvre2002;
@Ducatez2015;@Sol2005;@Overington2011;@HenkeVonDerMalsburg2020]. Finally, we 
assumed that shared phylogenetic history causes unobserved confounding and 
non-independence in unobserved tool use, feeding strategy, and relative brain 
size across the parrot phylogeny.

@@

Given our proposed causal model, we constructed a statistical model to impute 
unobserved probabilities of tool use and test existing theories of the evolution 
of tool use in parrots. To understand the model, suppose that we have the 
following observed variables for parrot species $i$. For the scientific 
literature, we declare $\text{N}_{\text{Lit},i}$ as the number of papers 
published before and up to tool use identification for species $i$ (or, if tool 
use has not been identified, the total number of papers published for species 
$i$) and $\text{T}_{\text{Lit},i}$ as a binary variable stating whether (1) or 
not (0) tool use has yet been observed in the scientific literature for species 
$i$. For the crowdsourced videos, we declare $\text{N}_{\text{Vid},i}$ as the 
number of videos published before and up to tool use identification for species 
$i$ (or, if tool use has not been identified, the total number of videos 
published for species $i$) and $\text{T}_{\text{Vid},i}$ as a binary variable 
stating whether (1) or not (0) tool use has yet been observed in the 
crowdsourced videos for species $i$. Additionally, $\text{F}_i$ and 
$\text{EQ}_i$ are feeding strategy and encephalisation quotient values for 
species $i$ and we have a phylogenetic distance matrix $D$ that describes the 
patristic distances between all parrot species.

>>>>

---
title             : "Digital video platforms and phylogenetic modelling reveal parrot tool use is not rare"
shorttitle        : "Parrot tool use"

@@

abstract: |
 Putatively rare behaviours, such as tool use, are challenging to study because absence of evidence can arise either from a species’ inability to produce the behaviour or from insufficient research effort. Here, we tackle this challenge by combining data from digital platforms and phylogenetic modelling to estimate actual rates of tool use in parrots. Videos on a social media platform revealed novel instances of tool use in 17 parrot species, more than doubling the confirmed number of tool-using parrot species from 11 (3%) to 28 (7%). Phylogenetic modelling ranked additional species that are most likely to be unobserved tool users, suggesting that between 11% and 17% of extant parrot species may be tool users. These discoveries have implications for inferences about the evolutionary drivers and origins of tool use in parrots, revealing associations with relative brain size and feeding generalism and indicating several genera where tool use was likely an ancestral trait. Overall, our findings challenge the assumption that current sampling effort captures the full distribution of putatively rare animal behaviours. Combining our sampling and analysis methods offers a fruitful approach for investigating the distribution, drivers, and origins of other rare behaviours.

keywords          : "tool use; technical intelligence; parrots; digital video platforms; phylogenetic modelling"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright † These authors contributed equally to this work \par ‡ These authors contributed equally to this work \par * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : current-biology.csl
---

# Introduction

@@

One potentially powerful method for quantifying actual rates of rare animal 
behaviours is by collating evidence of the behaviours from digital video 
platforms [@Nelson2013]. With millions of videos posted every day, digital 
platforms like YouTube offer a valuable source of data on animal behaviour. 
Digital video platforms have already been successfully used to uncover a variety
of rare animal behaviours, including interspecies play in dogs [@Nelson2013], 
novel problem-solving behaviours in horses [@Krueger2019], and death-related 
behaviours in Asian elephants [@Pokharel2022]. By casting the net wider than the 
scientific literature, digital video platforms can provide an indication of the 
tool-using species that the literature might be missing.

Even after collating evidence from digital video platforms, some tool-users 
could _still_ remain unobserved. One principled framework for identifying these
unobserved species is to specify a causal model of the process that generates
the observed data. We propose one such causal model in Figure 
\@ref(fig:plotDAG). In this model, we assume that the presence of tool use in 
the scientific literature (or on digital video platforms) is caused by both 
unobserved tool use capabilities and the number of published studies (or the 
number of videos) for any given species. Tool users are more likely to be 
observed if they are well studied, but understudied tool users may go 
undetected. Furthermore, based on existing theories of the evolution of tool use
[@Ducatez2015;@HenkeVonDerMalsburg2020;@Lefebvre1997;@Lefebvre2002;
@Lefebvre2004;@MettkeHofmann2014;@Overington2011;@Sol2005;@Sol2009], we propose
that the unobserved tool use capabilities are additionally caused by relative 
brain size, feeding strategy, and shared phylogenetic ancestry. Expressing this
causal model as a statistical model can suggest further species which are likely
to be unobserved tool-users and, simultaneously, test existing theories of the 
evolutionary drivers of tool use without incorrectly assuming that absence of 
evidence is evidence of absence.

(ref:plotDAGCaption) _Causal model of observed tool use._ Directed acyclic graph 
of the causal relationships between observed tool use and other variables. 
Available scientific data on tool use is caused both by unobserved tool use 
presence and scientific research effort (i.e., number of publications). 
Available video data on tool use is caused both by unobserved tool use 
presence and video research effort (i.e., number of videos). According to 
theory, unobserved tool use presence should be caused by relative brain size 
(encephalisation quotient) and feeding strategy (generalist vs. specialist). 
These variables all share unobserved confounds generated by shared phylogenetic
history. Grey circles indicate unobserved variables.

@@

Here, we apply these approaches to tool use in the parrot order. We focus on 
tool use in parrots for a number of reasons. First, the scientific literature 
suggests that only a small proportion of extant parrot species (11 out of 398;
3%) use tools [@Auersperg2009;@Auersperg2010;@Auersperg2011a;@Auersperg2011b;
@Auersperg2012;@Bastos2021;@BentleyCondit2010;@Borsari2005;@Goodman2018;
@Heinsohn2017;@Janzen1976;@Lambert2015;@Lefebvre2002;@OHara2021]. Parrot tool 
use thus provides an ideal test case for examining how robust sampling is in the 
scientific literature. Second, parrots are highly popular as pets. Over 70% of 
all extant parrot species are bred in the aviculture industry and kept as pets 
worldwide [@Anderson2003;@Carrete2008;@Drews2001;@Kelly2014;@Li2014;@Su2015;
@Young2012], enabling us to leverage the power of digital video platforms to 
search for evidence of tool use [@Nelson2013]. Third, detailed data on relative
brain sizes [@Flammer2001;@Iwaniuk2005;@Mazengenya2018;@Olkowicz2016;
@Schuck2008;@Silva2017], feeding strategies [@Wilman2016], and shared ancestry
[@Jetz2012] exist for parrots, allowing us to fit the statistical model implied
by Figure \@ref(fig:plotDAG) to the entire parrot order.

We first present the results from our video survey, in which we collated videos
of tool use in parrots from a digital video platform. This survey reveals a 
number of previously unidentified tool-using parrot species, which we map onto 
the phylogeny of the parrot order. We then describe our statistical model in 
more detail, and use it to (_i_) rank further parrot species that are likely 
unobserved tool users and (_ii_) re-examine key hypotheses regarding the 
evolutionary drivers and origins of tool use in parrots.

@@

## Digital video platform reveals tool use in additional parrot species

We surveyed the digital media platform YouTube for video evidence of tool use in 
parrots (see Methods for detailed search criteria). In our search, we used the 
standard criteria for identifying tool use in the literature, defining "true" 
tool use behaviour as the manipulation of an unattached object as an extension 
of the animal's body to achieve a goal [@Shumaker2011], while "borderline" tool 
use involved the use of an object that was still attached to a substrate
[@Seed2010].

@@

(ref:plotPhylo1Caption) _Results of video survey and phylogenetic survival cure
modelling mapped onto a maximum clade credibility phylogeny of the parrot order._
Orange points in the inner ring indicate species observed in the video survey,
with point size scaled by the number of videos for each species (note that three
species observed only in the video survey are not present in the phylogeny due
to a lack of genomic data: _Psittacara erythrogenys_, _Psittacus timneh_, and 
_Aratinga nenday_). Blue points in the outer ring indicate species observed in
the scientific literature, with point size scaled by the number of papers 
published on each species. Yellow species names indicate the top ten most likely
tool-using species from our phylogenetic survival cure model which were not 
observed in the scientific literature or the video survey. Total _n_ = 
`r nrow(tar_read(d))` species.

@@

In addition to the 28 tool-using species identified in the literature and our 
video survey, we fitted a Bayesian phylogenetic survival cure model to rank 
further species that are likely to be undetected tool-users (i.e., tool-using 
species with no tool use reported in the literature or in videos).

@@

Our tool use problem has the same features. We are modelling a time-to-event; 
specifically, the amount of "time" (i.e., observation opportunities measured as 
the number of published papers or videos) until tool use is identified. This is
right-censored data because many species will not have had tool use identified
when we measure them. Moreover, we can assume that a certain proportion of the
population will never experience the event -- that is, they are a non-tool-user
and so we will never identify tool use no matter how long we measure them for.

@@

Since the earliest anecdotes of parrots using tools by Wallace in the 1880s
[@Wallace1869] and more systematic anecdotal reports in the 1970s [@Goodall1971], 
only 11 parrot species (3% of all extant parrots) have been documented as tool 
users in the scientific literature. Our study collated data from a digital video
platform to identify 17 additional tool-using parrot species that are new to 
science, more than doubling the overall count to 28 species (7%). These species 
consisted of cockatoos (_Cacatuidae_), Old World parrots (_Psittacinae_), and 
neotropical parrots (_Arinae_). Beyond the video survey, the strong 
phylogenetic signal in our dataset allowed us to use phylogenetic information, 
along with other variables, to infer the unobserved probabilities of tool use 
across the parrot order. Our phylogenetic survival cure model incorporated
information on phylogenetic history, research effort, relative brain size, and 
feeding specialisation to rank parrot species that were most likely to be 
undetected tool users. The sum of probabilities from this model implied that 
between 15 and 41 of the species without observed evidence of tool use are 
likely to be undetected tool users, suggesting that the true proportion of tool
users may be as high as 17%.

@@

All instances of tool use in our video survey met the established criteria for 
tool use in the literature: the manipulation of an unattached object as an 
extension of the animal's body to achieve a goal [@Shumaker2011]. However, a 
striking feature of the dataset is that the 17 additional species identified in
the video survey were exclusively observed using tools for self-scratching.
While parrots in the wild use tools to achieve a variety of goals, including 
extractive foraging and preening [@Goodman2018;@OHara2021], parrots in captivity
only require tools to achieve the latter goal. Despite this difference in the
type of tool use observed, self-scratching is nonetheless a major category of 
animal tool use. Self-scratching also meets the definition for a more complex 
type of embodied tool use known as 'tooling', the deliberate generation of a 
mechanical interface by using an object to manipulate another target or surface
[@Fragaszy2018]. This makes it difficult to dismiss self-scratching as 
unimportant: to do so would require a redefinition of tool-use in non-human
animals. Moreover, the fact that we found strong phylogenetic clustering of 
self-scratching tool use from the video survey with other examples of tool use
from the literature supports a common underlying cognitive mechanism. In line 
with this, some of the species of parrots that use tools for self-scratching in
captivity also use tools for other purposes in the wild [@Auersperg2012;
@Bastos2021;@Goodman2018;@OHara2021].

There was little evidence to suggest that the observations of self-scratching
tool use in the video survey were merely unintentional accidents or explicitly 
trained behaviours. Individual parrots often used tools slowly and repetitively 
over long periods of time, even across multiple different videos, suggesting 
that their behaviour was not random or accidental [@Sandor2020]. Parrots 
preferentially employed self-scratching tools on areas of their body that were 
otherwise inaccessible, with 96% of all instances involving scratching of the 
head or neck, suggesting intentional tool use. For 60% of the species in the 
video survey, we found two or more videos of repetitive and sustained scratching 
by different individuals of the same species in separate households, suggesting 
that the manipulations were intentional and recurring events that did not 
represent unusual stereotypies of any single individual. Regarding the 
possibility of training or cueing from owners, over half of the videos contained 
no evidence of human interaction aside from filming the behaviour. Humans only 
handed parrots their tools in two of the videos, and none of the videos featured 
owners directly rewarding tool use behaviours with food. It is unlikely that 
owners would specifically train parrots to scratch themselves with tools. 
Finally, the high levels of phylogenetic signal in our data provide strong 
evidence that the observations from our video survey reflect
biologically-endowed capacities for tool use rather than accidental or trained
behaviours, which would likely appear uniformly across the phylogeny.

While missing data imputation is becoming more common in phylogenetic 
analyses [@Debastiani2021], the important distinction between absence of 
evidence and evidence of absence has not been given as much attention. Our 
phylogenetic analysis provides one approach to this problem by distinguishing 
between true absences of tool use and absences of tool use due to a lack of 
research effort in the scientific literature or in videos. To achieve this, we 
explicitly modelled the measurement of the outcome variable along a research 
effort time series, such that species with lower research effort in the 
literature or in videos were likely to be censored. In line with our causal 
model, we also included relative brain size, feeding strategy, and phylogenetic
history as predictors of unobserved tool use. We encourage researchers to test
this model by directing future study efforts towards the parrot species with the
highest probabilities of being undetected tool-users. Future work should also
refine the causal model in Figure \@ref(fig:plotDAG) to provide more certain
estimates of tool use probabilities, either by including additional predictor
variables or modelling further causes of measurement error in the taxonomic
record, such as species abundance and geographic accessibility [@Ducatez2014].

In conclusion, we have shown that the scientific literature has insufficiently 
captured the full distribution of tool use in the parrot order. Our 
digital video survey more than doubled the number of known tool using parrot 
species from 11 to 28, and our phylogenetic model suggested that the true 
proportion of parrot tool users could be as high as 17% of all species in this 
order. These discoveries have implications for theories of the evolutionary 
drivers and origins of tool use in parrots. Beyond parrot tool use, the 
methods used in this study have the potential to be applied to other rarely 
observed behaviours, including tool use in other taxa [@Hunt2013], rhythmic 
entrainment in birds [@Benichov2016;@Patel2009;@tenCate2016;@Wilson2016],
teasing behaviours in primates [@Eckert2020], and tactical deception across all
animals [@BroJorgensen2010;@Byrne1988;@Byrne1991]. We hope that these methods
will continue to uncover a diverse array of ephemeral behaviours that have as
yet gone undetected in the scientific literature.

@@

Our video search was conducted on YouTube in July 2020. Searches were conducted
manually by the first author over a month-long period, using the same IP address
and not logged in with a YouTube user account. Search terms included 
"parrot using tool" and variants (e.g., "macaw using tool", "lorikeet using 
tool", "parakeet using tool"), "tool use in parrot", "parrot tool use", "parrot 
scratching itself" (included after we found several videos demonstrating 
self-care tool use in previous searches) and equivalent terms (e.g., "parrot 
preening itself", "parrot grooming itself", "parrot scratching"). For all 
species that did not display results including object manipulation or scratching 
behaviours, we also searched the species’ common name(s) + "tool use", as well 
as the species’ common name(s) + "scratching". We also searched for translations 
of the terms "parrot tool use" and "parrot scratching" in languages for all 
countries where bird ownership was reported as >5% [@GlobalGfkSurvey], namely, 
Turkish, Czech, Polish, French, Italian, Dutch, German, Russian, Spanish, 
Portuguese, and Mandarin. Browser search histories were not cleared between
searches.

@@

Since YouTube is constantly growing and evolving in both its uploaded content 
and its proprietary recommendation algorithms, the same search strategy today 
would not return exactly the same videos as our search in July 2020. It is 
possible that some videos have since been deleted or more videos of parrot tool
use have since been uploaded. However, this does not affect our conclusions
about the number of parrot species found to use tools in our initial search.

@@

For modelling purposes, we coded research effort in both the scientific 
literature and the YouTube videos. For the scientific literature, we 
operationalised research effort as the number of papers published for each 
species' Latin name up to and including the first paper containing tool use for 
that species. If no tool use had been identified in the scientific literature 
for a species, then we coded the total number of papers published to date. We 
used the scientific database Scopus for coding the number of published papers. 
For the videos, we coded research effort as the number of search hits for each
species on YouTube. If tool use had been identified on YouTube, we estimated the
number of search hits when the first video of tool use was published on YouTube,
assuming linear growth of search hits since the inception of YouTube. If tool
use had not been identified, we used the current number of search hits.

@@

To infer unobserved probabilities of tool use across parrots, we proposed a 
causal model of observed tool use (Figure \@ref(fig:plotDAG)). We assumed that 
observed tool use in the scientific literature and in the videos is caused by 
both the unobserved presence or absence of tool use and research effort, proxied
by the number of papers published on a species and the number of videos
published on a species. Tool users are more likely to be observed if they are
well studied, but understudied tool users may go undetected. In addition, based
on theory, we also assumed that unobserved tool use is caused by feeding 
strategy and relative brain size [@Lefebvre2004;@Lefebvre1997;@Lefebvre2002;
@Ducatez2015;@Sol2005;@Overington2011;@HenkeVonDerMalsburg2020]. Finally, we 
assumed that shared phylogenetic history causes unobserved confounding and 
non-independence in unobserved tool use, feeding strategy, and relative brain 
size across the parrot phylogeny.

@@

Given our proposed causal model, we constructed a statistical model to impute 
unobserved probabilities of tool use and test existing theories of the evolution 
of tool use in parrots. To understand the model, suppose that we have the 
following observed variables for parrot species $i$. For the scientific 
literature, we declare $\text{N}_{\text{Lit},i}$ as the number of papers 
published before and up to tool use identification for species $i$ (or, if tool 
use has not been identified, the total number of papers published for species 
$i$) and $\text{T}_{\text{Lit},i}$ as a binary variable stating whether (1) or 
not (0) tool use has yet been observed in the scientific literature for species 
$i$. For the videos, we declare $\text{N}_{\text{Vid},i}$ as the 
number of videos published before and up to tool use identification for species 
$i$ (or, if tool use has not been identified, the total number of videos 
published for species $i$) and $\text{T}_{\text{Vid},i}$ as a binary variable 
stating whether (1) or not (0) tool use has yet been observed in the 
videos for species $i$. Additionally, $\text{F}_i$ and 
$\text{EQ}_i$ are feeding strategy and encephalisation quotient values for 
species $i$ and we have a phylogenetic distance matrix $D$ that describes the 
patristic distances between all parrot species.
